---
title: "Activity 2 (Ehall)"
author: "EMH"
date: "2025-10-30"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(gtsummary)
library(tidyverse)
CPAPAdherence_Data_Clean_1_ <- read_excel("/Applications/Biometry/Activity 2 WF/CPAPAdherence_Data_Clean (1).xlsx")
CPAPAdherence_Data_Clean_Dictionary_1_ <- read_excel("/Applications/Biometry/Activity 2 WF/CPAPAdherence_Data_Clean_Dictionary (1).xlsx")
```
# 1 This problem relies on the following set of variables: ethnicity, education level, race, age, sex, BMI, AHI, ESS, MMSE, adherence (for grouping), ODSI score at baseline, and ADCS-MCI caregiver score at 12 months.

A. Create a “Table 1: Demographic and Clinical Characteristics (n=174)” using the following notes: Column headers should include “Characteristic”, along with “Total”, “Non-adherent”, “Adherent”, and “Effect Size”. The last column should provide a p-value for appropriate statistical comparisons of distributions across adherent groups for each characteristic. Include sample sizes in parentheses in the column headers for the adherent and non-adherent groups. Populate the cells with the appropriate measures of central tendency and variability and label appropriately. Include 95% confidence intervals where appropriate and label appropriately. Use footnotes to illustrate the statistical tests used for comparisons.

B. Choose a single characteristic with a significant p-value when comparing between adherence and non-adherent groups, and describe in 2-3 sentences what this means in plain English.
C. Choose a single characteristic with a non-significant p-value when comparing between adherence and non-adherent groups, and describe in 2-3 sentences what this means in plain English.





#2 In the Observation and Interview Based Diurnal Sleepiness Inventory (ODSI), each of the three items is rated on a seven-point Likert scale. The first item examines sleepiness during basic activities of daily living; the second item relates to falling asleep during periods of inactivity; and the third item asks about hours of daytime sleep. The total ODSI score ranges from 0 (no somnolence) to 24 (excessive somnolence), so a lower score is better. Test the null hypothesis that there is no difference in change from baseline to 6 months for ODSI for adherent versus non-adherent participants. Write out each step of the hypothesis test and clearly interpret your results in plain English in 2-3 sentences.



#3 Based on the ODSI, a cutoff score of 6 or higher is used to identify older adults with excessive daytime sleepiness. Test the null hypothesis that the probability of excessive daytime sleepiness is the same at baseline versus at 6 months. Write out each step of the hypothesis test and clearly interpret your results in plain English in 2-3 sentences.



#4 The Mini Mental State Examination (MMSE) is a tool that can be used to systematically assess mental status. It is an 11-question measure that tests five areas of cognitive function: orientation, registration, attention and calculation, recall, and language. The maximum score is 30, where higher scores are better. A score of 23 or lower is indicative of cognitive impairment.

A. Test the null hypothesis that the underlying population representing all study participants is not cognitively impaired, assuming a known population standard deviation equal to 2.0. Interpret the results in 2-3 sentences for a non-statistical audience.

B. Test the null hypothesis that the underlying population representing all study participants is not cognitively impaired, assuming the population standard deviation is unknown. Interpret the results in 2-3 sentences for a non-statistical audience.



#5 This problem relies on the variable representing average daily CPAP use.
A. Compute and report the mean and standard deviation for the variable representing average daily CPAP use.

B. Assuming the estimates you computed represent population parameters, estimate and report the proportion of participants with average daily CPAP use less than 3 hours.

C. In 2-3 sentences, explain the meaning of this probability in plain language suitable for a non-statistical audience.


#6 

#### Compute and report the mean and standard deviation for BMI.Assuming the estimates you computed represent population parameters, estimate and report the proportion of participants who are considered obese (BMI 30+). In 2-3 sentences, explain the meaning of this probability in plain language suitable for a non-statistical audience.
```{r BMI,echo=FALSE}
# Compute and report the mean and standard deviation for BMI
mean_bmi <- mean(CPAPAdherence_Data_Clean_1_$bmi, na.rm = TRUE)
sd_bmi <- sd(CPAPAdherence_Data_Clean_1_$bmi, na.rm = TRUE)

mean_bmi
sd_bmi

# Estimate and report the proportion of participants with BMI 30+
obese <- mean(CPAPAdherence_Data_Clean_1_$bmi >= 30, na.rm = TRUE)
obesity_percentage <- obese * 100
print(obesity_percentage)
```


##### About 96% of participants have a body mass index of 30 or higher. This classifies 96% or 96 people out of every group of 100 people as obese.




#7 This problem relies on the Apnea Hypopnea Index (AHI) variable. Use the SamplingScript.R Download SamplingScript.Rfor drawing random samples.
A. Among those who are adherent to CPAP, describe the distribution of the variable AHI with appropriate statistics and data visualizations. Then, summarize your findings in 2-3 sentences in plain language suitable for a non-statistical audience.

B. Among those who are adherent to CPAP, describe the sampling distribution based on samples of size 30 from the variable AHI based on theory with appropriate statistics and data visualizations. Then, summarize findings in 2-3 sentences in plain language suitable for a non-statistical audience.

C. Randomly draw 1,000 samples of size 30 (with replacement) from the 128 AHI scores of adherent participants. For each sample, compute the sample mean AHI. Use these 1,000 sample means to:

i. Calculate and report the mean and standard deviation of this sampling distribution.

ii. Plot the sampling distribution of the sample means using a histogram.

iii. Overlay a normal curve on your histogram of sample means.

iv. In 1-2 sentences, describe how your sampling distribution compares to a normal distribution. Summarize the shape and spread of the sampling distribution in plain language.

v. In 2-3 sentences, describe how this distribution compares to your response from part (b) in plain language.

D. Among those who are non-adherent to CPAP, describe the sampling distribution based on samples of size 100 from the variable AHI based on theory with appropriate statistics and data visualizations. Then summarize your findings in 2-3 sentences in plain language suitable for a non-statistical audience.

E. Randomly draw 1,000 samples of size 100 (with replacement) from the 46 AHI scores of non-adherent participants. For each sample, compute the sample mean AHI. Use these 1,000 sample means to:

i. Calculate and report the mean and standard deviation of this sampling distribution.

ii. Plot the sampling distribution of the sample means using a histogram.

iii. Overlay a normal curve on your histogram of sample means.

iv. In 1-2 sentences, describe how your sampling distribution compares to a normal distribution. Summarize the shape and spread of the sampling distribution in plain language.

v. In 2-3 sentences, describe how this distribution compares to your response from part (d) in plain language.

F. Compare the sampling distributions among adherent versus non-adherent participants. How do the means and spreads of the two sampling distributions compare? Why do you think they are the same or different? Summarize your findings in 3-5 sentences in plain language suitable for a non-statistical audience.
